---
title: "wildfire_trend_analysis"
output: html_document
date: "2024-04-07"
---

#Initial analysis of wildifre trends from dataset
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')
```

* [Load Data](#Load)
* [Wildfires over Time](#Time)
* [Fires by Size](#Size)
* [Wildfire Causes](#Causes)
* [Wildfires by Geography](#Geo)

load libraries

```{r, message = FALSE}
library(RSQLite)
library(dbplyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(xts)
library(ggfortify)
library(ggthemes)
library(maps)
library(mapdata)
library(leaflet)
library(remotes)
library(qst)
library(ymd)
library(tidyverse)
```

#Create database connection, load downloaded dataset from kaggle (sqlite) into a dataframe
```{r}

# create db connection
conn <- dbConnect(SQLite(), 'D:/Big Data Architecture/FPA_FOD_20170508.sqlite')

# pull the fires table into RAM
fires <- tbl(conn, "Fires") %>% collect()

# check size
print(object.size(fires), units = 'Gb')

# disconnect from db
dbDisconnect(conn)

#write database as a csv
write.csv(fires, file = "wildfires.csv")

#take a look at the data
glimpse(fires)

```

#Analyze some basic trends. Look at fires by year
```{r}

fires %>% 
    group_by(FIRE_YEAR) %>%
    summarize(n_fires = n()) %>%
    ggplot(aes(x = FIRE_YEAR, y = n_fires/1000)) + 
    geom_bar(stat = 'identity', fill = 'orange') +
    geom_smooth(method = 'lm', se = FALSE, linetype = 'dashed', size = 0.4, color = 'red') + 
    labs(x = '', y = 'Number of wildfires (thousands)', title = 'US Wildfires by Year')

```

Spike in fires around 2006, slight general trend of increasing amount of fires over time

#Analyze trends by day of year (to see during which seasons there are most fires)
```{r}

fires %>% 
    group_by(DISCOVERY_DOY) %>%
    summarize(n_fires = n()) %>%
    ggplot(aes(x = DISCOVERY_DOY, y = n_fires)) + 
    geom_line(color = 'orange') +
    geom_smooth(method = 'lm', se = FALSE, linetype = 'dashed', size = 0.4, color = 'red') + 
    labs(x = '', y = 'Number of wildfires', title = 'US Wildfires by Day of Year')

```

#Create time-series of by-day fire for total timespan of dataset
```{r}

fires$CONT_DATEymd <- as.Date(fires$CONT_DATE - 2458014.5, origin = '2017-09-18')
fires$DISCOVERY_DATEymd <- as.Date(fires$DISCOVERY_DATE - 2458014.5, origin = '2017-09-18')

fires %>% 
    group_by(DISCOVERY_DATEymd) %>%
    summarize(n_fires = n()) %>%
    ggplot(aes(x = DISCOVERY_DATEymd, y = n_fires)) +
    geom_line(color = 'orange') +
    labs(x= '', y = 'Number of fires', title = 'Daily number of fires 1992-2015')

```

#Simplify by-day timespan into monthly timespan
```{r}

fires_mo <- fires %>% 
    group_by(DISCOVERY_DATEymd) %>%
    summarize(n_fires = n())

fires_mo <- as.xts(fires_mo$n_fires, order.by = fires_mo$DISCOVERY_DATEymd) %>%
    apply.monthly(FUN = sum)

autoplot(fires_mo, ts.colour = 'orange') +
    labs(y = 'Number of fires', title = 'Monthly number of fires 1992-2015')

```

#Analyze fires by size. Divide dataset into fire size classes
```{r}

size_classes <- c('A' = '0-0.25', 'B' = '0.26-9.9', 'C' = '10.0-99.9', 'D' = '100-299', 'E' = '300-999',
                  'F' = '1000-4999', 'G' = '5000+')

fires %>% 
    group_by(FIRE_SIZE_CLASS) %>%
    summarize(n = n()) %>%
    mutate(FIRE_SIZE_CLASS = size_classes[FIRE_SIZE_CLASS]) %>%
    ggplot(aes(x = FIRE_SIZE_CLASS, y= n)) +
    geom_bar(stat = 'identity', fill = 'Orange') +
    labs(x = 'Fire size (acres)', y = 'Number of fires', title = 'Number of Wildfires by Size Class')

```

#Analyze the causes of fires
```{r}

fires %>%
    group_by(STAT_CAUSE_DESCR) %>%
    summarize(n_fires = n()/1000) %>%
    ggplot(aes(x = reorder(STAT_CAUSE_DESCR, n_fires), y = n_fires)) +
    geom_bar(stat = 'identity', fill = 'orange') + 
    coord_flip() + 
    labs(x = '', y = 'Number of fires (thousands)', title = 'US Wildfires by Cause 1992 to 2015')
    
```

#Try to find which cause of fire resulted in the largest fires by area burned
```{r}

fires %>% 
    group_by(STAT_CAUSE_DESCR) %>%
    summarize(mean_size = mean(FIRE_SIZE, na.rm = TRUE)) %>%
    ggplot(aes(x = reorder(STAT_CAUSE_DESCR, mean_size), y = mean_size)) +
    geom_bar(stat = 'identity', fill = 'orange') + 
    coord_flip() + 
    labs(x = '', y = 'Acres', title = 'Average Wildfire Size by Cause')
```

#Analyze wildfires by state. See which states had the most
```{r}

# Add codes for DC and Puerto Rico to the default state lists
state.abb <- append(state.abb, c("DC", "PR"))
state.name <- append(state.name, c("District of Columbia", "Puerto Rico"))

# Map the stat1e abbreviations to state names so we can join with the map data
fires$region <- map_chr(fires$STATE, function(x) { tolower(state.name[grep(x, state.abb)]) })

# Get the us state map data
state_map <- map_data('state')

```


```{r}

fires %>% 
    select(region) %>%
    group_by(region) %>%
    summarize(n = n()) %>%
    right_join(state_map, by = 'region') %>%
    ggplot(aes(x = long, y = lat, group = group, fill = n)) + 
    geom_polygon() + 
    geom_path(color = 'white') + 
    scale_fill_continuous(low = "orange", 
                          high = "darkred",
                          name = 'Number of fires') + 
    theme_map() + 
    coord_map('albers', lat0=30, lat1=40) + 
    ggtitle("US Wildfires, 1992-2015") + 
    theme(plot.title = element_text(hjust = 0.5))

#most fires in California and Texas, surprisingly a lot in Georgia
```
